---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visualization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(RColorBrewer)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTuesday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2020/2020-09-22/expeditions.csv", sep = ',', stringsAsFactors = FALSE) # Reading the raw dataset from link
peak_times <- expeditions %>%
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>%  # Keep only the top 15 peaks, move the rest into "Other"
  filter(peak_name != "Other") %>%  # Remove the "Other" category
  count(peak_name, season) %>%  # Grouping by name of the peak and season and counting the rows
  group_by(peak_name) %>% # Group these by the name of the peak
  mutate(sum_expeditions = sum(n)) %>%  # Calculate total expeditions per peak
  ungroup() %>% # Remove grouping because of further data manipulation
  mutate(peak_name = fct_reorder(peak_name, sum_expeditions, .desc = FALSE)) # Reordering the factors by number of expeditions per peak

ggplot(peak_times, aes(x = n, y = peak_name, fill = season)) +
  geom_bar(stat = "identity") +  # Bar chart with width proportional to total expeditions
  scale_fill_viridis_d(option = "D") +  # Apply viridis palette with default colors
  theme_light() + # Light theme
  theme(axis.text.x = element_text(angle = 0, hjust = 1), legend.position = "bottom") + # Position and angle of texts
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    x = "Number of expeditions",
    y = "",
    fill = "season"
  )# Captions
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTuesday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phds <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-02-19/phd_by_field.csv", sep = ',', stringsAsFactors = FALSE) # Reading the raw dataset from link
sum_phds_per_fields <- phds %>%
  group_by(year, broad_field) %>% # Group the data by year and field
  summarise(sum_phds = sum(n_phds, na.rm = TRUE), .groups = "drop") # Count the total number of PhD-s by the groups

ggplot(sum_phds_per_fields, aes(x = year, y = sum_phds, color = broad_field)) +
  geom_line(linewidth = 1.2) + # Line plot with size = 1.2
  scale_fill_brewer(palette = "Dark2") + # Colors from Dark2 palette
  scale_y_continuous(labels = comma_format()) + # Comma-formatted labels on Y-axis
  scale_x_continuous(breaks = pretty_breaks()) + # Pretty breaks on X-axis
  labs(title = "Number of awarded Ph.D.-s in the US by year", x = "", y = "", color = "Broad field") + # Captions
  theme_minimal() + # Minimalistic theme
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) # Angle of elements
```

## Task 3: Commute in the US

The 2019-11-05 TidyTuesday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commutes <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-11-05/commute.csv", sep = ',', stringsAsFactors = FALSE) # Reading the raw dataset from link
walk_data <- commutes %>% 
  filter(mode == "Walk") %>% # First we filter the dataset of people walking
  group_by(state, state_region, state_abb) %>% # Group the dataset by state, state region and the abbreviation
  summarise(sum_walk = sum(n, na.rm = TRUE), .groups = "drop") # Count the number of rows and drop the grouping
bike_data <- commutes %>%
  filter(mode == "Bike") %>% # Next we filter the dataset of people biking
  group_by(state, state_region) %>% # Group the dataset by state, state region
  summarise(sum_bike = sum(n, na.rm = TRUE), .groups = "drop") # Count the number of rows and drop the grouping

commutes_used <- left_join(walk_data, bike_data, by = "state") %>% # Merging the two datasets (because of merging by state, state region will be duplicated)
  select(-state_region.y) %>% # Remove the duplicate column
  rename(state_region = state_region.x) # Rename the remaining column
  
ggplot(commutes_used, aes(x = sum_walk, y = sum_bike, color=state_region)) +
  geom_point(size = 2) +  # Scatter plot with size = 2
  geom_text(aes(label = ifelse(!is.na(state_abb), state_abb, "")), color = "black", size = 5) + # Label all points with the abbreviations if the value is NOT NA
  scale_x_log10(labels = comma_format()) +  # Log-transformed x-axis with comma-formatted labels
  scale_y_log10(limits = c(30, NA), labels = comma_format()) +  # Log-transformed y-axis with comma-formatted labels and the low limit of value = 30
  labs(title = "Title number of people walking vs. biking to work in each USA state", x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", color = "State region") + # Captions
  theme_light() +  # Light theme
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) # Angle and positioning of text
```
