---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visualization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(dplyr)
library(ggplot2)
library(forcats)
library(scales)
library(RColorBrewer)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTuesday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions_data <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2020/2020-09-22/expeditions.csv", sep = ',', stringsAsFactors = FALSE)
top_peaks <- expeditions_data %>%
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>%  # Keep only the top 15 peaks, lump others into "Other"
  filter(peak_name != "Other") %>%  # Remove the "Other" category
  count(peak_name, season) %>%  # Count expeditions per peak and season
  group_by(peak_name) %>% 
  mutate(total_expeditions = sum(n)) %>%  # Calculate total expeditions per peak
  ungroup() %>%
  mutate(peak_name = fct_reorder(peak_name, total_expeditions, .desc = FALSE)) 

ggplot(top_peaks, aes(x = n, y = peak_name, fill = season)) +
  geom_bar(stat = "identity") +  # Bar chart with height proportional to expeditions_count
  scale_fill_viridis_d(option = "D") +  # Apply viridis palette
  theme_minimal() +  # Use a light theme
  theme(axis.text.x = element_text(angle = 0, hjust = 1), legend.position = "bottom") +  # Rotate x-axis text
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    x = "Number of expeditions",
    y = "",
    fill = "season"
  )
  
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_data <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-02-19/phd_by_field.csv", sep = ',', stringsAsFactors = FALSE)
phd_aggregated <- phd_data %>%
  group_by(year, broad_field) %>%
  summarise(total_phds = sum(n_phds, na.rm = TRUE)) %>%
  ungroup

ggplot(phd_aggregated, aes(x = year, y = total_phds, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_fill_brewer(palette = "Dark2") +
  scale_y_continuous(labels = comma_format()) +
  scale_x_continuous(breaks = pretty_breaks()) +
  labs(title = "Number of awarded Ph.D.-s in the US by year", x = "", y = "", color = "Broad field") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 0, hjust = 1))
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute_data <- read.csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/refs/heads/main/data/2019/2019-11-05/commute.csv", sep = ',', stringsAsFactors = FALSE)
walk_aggregated <- commute_data %>%
  filter(mode == "Walk") %>%
  group_by(state, state_region, state_abb) %>%
  summarise(total_walk_n = sum(n, na.rm = TRUE))
bike_aggregated <- commute_data %>%
  filter(mode == "Bike") %>%
  group_by(state, state_region) %>%
  summarise(total_bike_n = sum(n, na.rm = TRUE))

walk_bike_merge <- left_join(walk_aggregated, bike_aggregated, by = "state")
  
ggplot(walk_bike_merge, aes(x = total_walk_n, y = total_bike_n, color=state_region.x)) +
  geom_point(size = 2) +  # Plot points with size = 2
  geom_text(aes(label = state_abb), color = "black", size = 5) +
  scale_x_log10(labels = comma_format()) +  # Log-transformed x-axis with comma-formatted labels
  scale_y_log10(limits = c(30, NA), labels = comma_format()) +  # Log-transformed y-axis with comma-formatted labels
  labs(title = "Title number of people walking vs. biking to work in each USA state", x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", color = "State region") +
  theme_light() +  # Light theme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```
